# Intro to ggplot

We’ll use the **ICAN-ICAR 2025** survey data with variables such as:

-   `ReadingIRTScore`: Child's reading latent ability score
-   `MathIRTScore`: Child's maths latent ability score
-   `ch02`: Child's age
-   `ch03`: Child's gender
-   `EnrolmentStatus`: child's school enrollment status
-   `ch04a`: whether the child has eye difficulties

Load the packages and load and prepare the data:

```{r}
#| echo: false
#| warning: false
#| message: false

# Load packages
library(tidyverse)
library(survey)
library(broom)
library(knitr)


# Load the data
dat = read_csv("data/ican-icar-2025-v1.csv") |>
      mutate(ch03 = factor(ch03, levels = c(1, 2), labels = c("Female", "Male"))) |>
      filter(CountryName == "Senegal") |> 
      select(ReadingIRTScore, MathIRTScore,ch03, ch02, ch04a, EnrolmentStatus) 

```

You don’t need to understand all the details of `ican-icar-2025-v1` yet.\
For now, just remember:

-   We’ll be using a data frame called **`dat`**.
-   Each **row** is a child.

------------------------------------------------------------------------

## What is ggplot2?

`ggplot2` is the R package we’ll use to make graphs.

-   It’s part of the **tidyverse** (a family of R packages for data analysis).
-   It’s based on the **Grammar of Graphics** idea.
-   Instead of giving you a few pre-made plots, it lets you build plots from **simple pieces** (layers).
-   You can use it effectively **without** fully understanding the underlying theory (we’ll learn by doing).

In practice, we’ll keep this simple mental model:

> **To make a plot in ggplot2 we always say:**\
> “Use this **data**, map these **variables** to the axes/colour/etc.,\
> and draw them with this **geom** (points, bars, lines, …).”

### Optional: Grammar of Graphics (background)

If you’re curious about the theory:

-   **Graphics = distinct layers** (data, aesthetic mappings, geoms, …).
-   **Aesthetic mapping**: connect variables in the data (numbers, labels) to what we see (position, colour, size, shape).

Don’t worry if this feels abstract right now; the examples will make it concrete.

------------------------------------------------------------------------

## The main pieces of ggplot2

A **ggplot2** plot is built from several elements.

For our intro, we only need these three:

1.  **data** – the data frame we’re plotting (here: `dat`).
2.  **aes(...)** – *aesthetic mappings*: how variables map to what we see (x-axis, y-axis, colour, size, shape).
3.  **geom\_…()** – the *geometric object* that draws the data (points, lines, bars, etc.).

Later (optional), we can also use:

-   **stats** – automatic summaries (e.g. means, counts, smooth lines).
-   **scales** – axis ranges and colour scales.
-   **coordinate systems** – how axes are drawn.
-   **facets** – small multiples of the same plot.
-   **themes** – fonts, grid lines, background (“non-data ink”).

------------------------------------------------------------------------

## Building your first plot: data → aes → geom

We’ll start by plotting **household expenditure** against **household income**.

### Step 1: Choose the data

First we tell ggplot2 *which data frame to use*:

```{r}
ggplot(data = dat)
```

This creates a **blank plotting area** (a coordinate system), but we haven’t told it *what* to draw yet, so you’ll likely see an empty plot.

### Step 2: Map variables with `aes()`

Next, we say which variables go to which axis using `aes()`:

```{r}
ggplot(data = dat,
       mapping = aes(x = ReadingIRTScore, y = MathIRTScore))
```

Read this as:

-   Put `ReadingIRTScore` on the **x-axis** (horizontal).
-   Put `ch02` on the **y-axis** (vertical).

This still doesn’t draw points — it only defines the **mapping**.

### Step 3: Add a geometry with `geom_point()`

Now we add a **geometry** to actually draw the data. For a scatterplot, we use `geom_point()`:

```{r}
ggplot(data = dat,
       mapping = aes(x = ReadingIRTScore, y = MathIRTScore)) +
  geom_point()
```

Think of this as:

> “Using data `dat`,\
> put reading score on x, maths score on y,\
> and draw one **point per child**.”

This **data + aes + geom** pattern is the core template you’ll reuse for almost every plot in ggplot2. You can find other geometries that might interest you, i.e., a bar graph is `geom_bar()`, a line graph is `geom_line()`, etc. Here, we keep it simple by honing a scatter plot to show many design controls you could attain from **ggplot**.

------------------------------------------------------------------------

## Aesthetics: colour, shape, size

`aes()` can also control things like **colour**, **shape**, and **size**.

Big idea:

> Whatever you put *inside* `aes(...)` is controlled by the **data**.\
> For example, `aes(colour = ch03)` means “use the variable `ch03` to decide the colour of each point”.

We’ll reuse our scatterplot and experiment.

### Mapping shape to a variable

```{r}
ggplot(data = dat,
       mapping = aes(x = ReadingIRTScore, 
                     y = MathIRTScore, 
                     shape = ch03)) +
  geom_point()
```

-   Different gender groups (`ch03`) get different **shapes**.

### Mapping colour to a variable

```{r}
ggplot(data = dat,
       mapping = aes(x = ReadingIRTScore, 
                     y = MathIRTScore, 
                     color = ch03)) +
  geom_point()
```

-   Different groups get different **colours**.
-   A **legend** is added automatically.

### Mapping size to a variable (e.g., age)

```{r}
ggplot(data = dat,
       mapping = aes(x = ReadingIRTScore, 
                     y = MathIRTScore, 
                     size = ch02)) +
  geom_point(alpha = 0.1)
```

-   Children with higher `ch02` get **larger points**.
-   `alpha = 0.1` makes points more transparent, so dense regions are easier to see.

### Using a categorical variable

If you have categories like `EnrolmentStatus`, you can map those to colour or fill:

```{r}
dat = dat |>
       filter(!is.na(ReadingIRTScore), !is.na(MathIRTScore))
ggplot(data = dat,
       mapping = aes(x = ReadingIRTScore,
                     y = MathIRTScore,
                     colour = EnrolmentStatus)) +
  geom_point(alpha = 0.25)
```

-   Each BMI category gets a different **colour**.
-   Missing categories are filtered out with `filter(!is.na(bmi.bins))`.

------------------------------------------------------------------------

## Aesthetic mapping vs fixed settings (some nuance)

Sometimes you want a fixed colour for all points (not data-driven).

-   **Inside `aes()`** → the value comes from the data.
-   **Outside `aes()`** → you set a fixed value.

For example:

```{r}
# Colour determined by data (race)
ggplot(data = dat,
       mapping = aes(x = ReadingIRTScore,
                     y = MathIRTScore,
                     colour = EnrolmentStatus)) +
  geom_point(alpha = 0.25)

# Fixed colour (all points blue)
ggplot(data = dat,
       mapping = aes(x = ReadingIRTScore,
                     y = MathIRTScore,
                     colour = EnrolmentStatus)) +
  geom_point(alpha = 0.25, colour = "steelblue")
```

This distinction becomes important when you combine multiple layers or want to control colours manually, but for now just note the pattern.

------------------------------------------------------------------------

## Small polish: labels and a simple theme

Let’s tidy up our core scatterplot a little bit:

```{r}
library(scales)

ggplot(data = dat,
       mapping = aes(x = ReadingIRTScore,
                     y = MathIRTScore,
                     colour = EnrolmentStatus)) +
  geom_point(alpha = 0.25) +
  labs(
    x = "Reading latent ability score",
    y = "Maths latent ability score",
    colour = "School enrollment status",
    title = "Foundational Skills in Senegal",
    subtitle = "ICAN-ICAR 2025 Survey"
  ) +
  scale_x_continuous(labels = label_comma()) +
  scale_y_continuous(labels = label_comma()) +
  theme_classic()
```

Here we:

-   Use `labs()` to set **axis labels**, **legend title**, and **plot title/subtitle**.
-   Use `theme_minimal()` to clean up the background and gridlines.
-   Just *tip*: by default, ggplot sometimes shows big numbers in scientific notation. We can tell ggplot to use "normal looking" numbers with commas by adding **scale_x_continuous()** etc.

> For a intro, this level of theme usage is enough.\

------------------------------------------------------------------------

## Practice: Test your knowledge with `mpg`

The `ggplot2` package comes with a built-in dataset called `mpg`.

Load it:

```{r}
library(ggplot2)  # already loaded via tidyverse, but explicit is fine
library(skimr)

data(mpg)

# Optional: look at its structure
skim(mpg)
```

You can use `?mpg` or `help(mpg)` to see more information about the variables.

Using the template:

``` r
ggplot(data = <DATA>,
       mapping = aes(x = <X>, y = <Y>, ...)) +
  <GEOM_FUNCTION>()
```

try these:

1.  Use a **scatterplot** to show the relationship between `displ` (engine displacement, in litres) and `hwy` (highway miles per gallon)
2.  In the same plot, set the **colour of the points** to `class`
3.  Graph a **boxplot** of `hwy` by `class`.
4.  Graph a **bar plot** of `class`.

------------------------------------------------------------------------

## Saving plots to files with `ggsave()`

You can save the last plot as a PDF:

```{r}
# Save last displayed plot as PDF
ggsave("figs/myggplot.pdf")
```

Or save a named plot object:

```{r}
p = ggplot(data = dat,
       mapping = aes(x = ReadingIRTScore,
                     y = MathIRTScore,
                     colour = EnrolmentStatus)) +
  geom_point(alpha = 0.25) +
  labs(
    x = "Reading latent ability score",
    y = "Maths latent ability score",
    colour = "School enrollment status",
    title = "Foundational Skills in Senegal",
    subtitle = "ICAN-ICAR 2025 Survey"
  ) +
  scale_x_continuous(labels = label_comma()) +
  scale_y_continuous(labels = label_comma()) +
  theme_classic()

ggsave(filename = "figs/myggplot.png",
       plot   = p,
       width  = 10,
       height = 10,
       units  = "in")
```

-   The file type is guessed from the extension (`.pdf`, `.png`, etc.).
-   You can control size with `width`, `height`, and `units`.

------------------------------------------------------------------------

##  Useful resources

Some excellent follow-up resources:

-   [ggplot2 documentation](https://ggplot2.tidyverse.org/) (function reference is especially useful).\
-   [*R for Data Science*](https://r4ds.hadley.nz/) – Data Visualization chapter.\
-   RStudio’s [ggplot2 cheatsheet](https://rstudio.github.io/cheatsheets/data-visualization.pdf).\
-   [R Graphics Cookbook](https://osctr.ouhsc.edu/sites/default/files/2020-02/rcourse/3/RGraphicsCookbook.pdf) (for a recipe-style approach).
-   Asano Masahiko has cool section on [Data Visualization](https://www.asanoucla.com/introduction-to-r/) using ggplot.

```{r}
# Friendly reminder: once you're comfortable with the basics,
# pick one plot you make often and try to recreate it using ggplot2.It can be a bar graph, histogram, or a box plot

```
