# Simple Linear Regression

Regression analysis describes how an outcome changes as a predictor changes. In this chapter, you will learn how to summarise the association between two numeric variables with a correlation coefficient and how to estimate a simple linear regression model. Because ICAN-ICAR uses a complex survey design, you will also fit the same model with survey weights, clustering, and stratification so that inference reflects the sampling design.

```{r}
#| echo: false
#| warning: false
#| message: false

# Load packages
library(tidyverse)
library(survey)
library(broom)
library(knitr)

# Load the data
dat = read_csv("data/ican-icar-2025-v1.csv")

```

## Preparing variables for analysis

We study the relationship between a child’s grade (`ch06a`) and a literacy score constructed from item-level responses. Grade is a plausible predictor because it often reflects additional exposure to schooling. Regression can quantify association, but it does not by itself establish causality; use theory and study design to justify causal claims.

**Literacy score.** The literacy items are coded so that `2` indicates a correct response and `0` or `1` indicates an incorrect response. We recode correct responses to `1`, recode incorrect responses to `0`, and sum across items. This scoring treats missing item responses as incorrect when it computes totals (`na.rm = TRUE`). Use a different rule if your assessment protocol defines missingness differently.

```{r}
#| label: create literacy score
#| warning: false
#| message: false

# grade (keep plausible values)
dat = dat |>
  mutate(
    grade = ifelse(ch06a >= 0 & ch06a <= 12, ch06a, NA_real_)
  )

# literacy items
literacy_items = c(
  paste0("l1_", 1:4),
  paste0("l2_", 1:5),
  paste0("l3_", 1:5),
  paste0("l4_", 1:6),
  paste0("l5_", 1:5),
  paste0("l6_", 1:5)
)

dat = dat |>
  mutate(
    across(
      all_of(literacy_items),
      ~ case_when(
        .x == 2 ~ 1,
        .x %in% c(0, 1) ~ 0,
        TRUE ~ NA_real_
      )
    ),
    literacy_score = rowSums(across(all_of(literacy_items)), na.rm = TRUE)
  )
```

For the examples below, we work with one country to keep output readable. Replace the value of `country_name` to analyse a different country.

```{r}
#| label: analysis sample
#| warning: false
#| message: false

country_name = "Kenya"

d = dat |>
  filter(
    CountryName == country_name,
    Location == "Urban",       # subset to only urban areas
    !is.na(grade),
    !is.na(literacy_score)
  )
```

## Correlation and simple regression

Correlation provides a first summary of a linear relationship between two numeric variables. A correlation near `0` indicates a weak linear association, while values close to `1` or `-1` indicate a strong positive or negative linear association. Because correlation targets linear patterns, treat it as a starting point rather than a complete description.

```{r}
#| label: unweighted correlation
#| warning: false
#| message: false

cor_unweighted = cor(d$grade, d$literacy_score, use = "complete.obs")
cor_unweighted
```

**A quick check.** A scatterplot helps you assess whether a linear relationship is plausible and whether a small number of extreme values dominate the pattern.

```{r}
#| label: scatterplot
#| warning: false
#| message: false

ggplot(d, aes(x = grade, y = literacy_score)) +
  geom_point(alpha = 0.35) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    x = "Grade",
    y = "Literacy score",
    title = "Literacy score and grade in urban Kenya"
  ) +
  theme_bw()
```

A simple linear regression model has the form

\[
Y_i = \alpha + \beta X_i + \varepsilon_i,
\]


where \(\beta\) (the slope) describes the expected change in \(Y\) for a one-unit increase in \(X\). In our example,

\[
\text{literacy\_score}_i = \alpha + \beta \cdot \text{grade}_i + \varepsilon_i.
\]

The unweighted model describes the relationship in the sample.

```{r}
#| label: lm unweighted
#| warning: false
#| message: false

m_lm = lm(literacy_score ~ grade, data = d)

tidy(m_lm, conf.int = TRUE) 

```

## Survey-weighted regression for population inference

To make population statements, fit the regression with the survey design. This approach accounts for weights, clustering, and stratification when it computes standard errors.

```{r}
#| label: survey design object
#| warning: false
#| message: false

options(
  survey.lonely.psu = "adjust",
  survey.adjust.domain.lonely = TRUE
)

des = svydesign(
  ids     = ~interaction(CountryName, VillageID) + HHID,
  strata  = ~interaction(CountryName, TierOneUnit),
  weights = ~HHWeightProvided,
  data    = dat,
  nest    = TRUE
)

des_country = subset(des, CountryName == country_name & !is.na(grade) & !is.na(literacy_score))
```

```{r}
#| label: weighted correlation
#| warning: false
#| message: false

V = svyvar(~grade + literacy_score, design = des_country, na.rm = TRUE)

cor_weighted = V[1, 2] / sqrt(V[1, 1] * V[2, 2])
cor_weighted
```

```{r}
#| label: svyglm weighted
#| warning: false
#| message: false

m_svy = svyglm(literacy_score ~ grade, design = des_country)

tidy(m_svy, conf.int = TRUE) 
```

Interpret the slope as the expected change in the literacy score associated with a one-grade increase, holding the model form fixed. Use the survey-based standard errors and confidence intervals for inference.

**Predicted values.** Predictions translate the fitted line into an expected score at a specific grade. The example below reports the predicted literacy score for grade 3, with a standard error from the survey model.

```{r}
#| label: prediction grade 3
#| warning: false
#| message: false

new = tibble(grade = 3)

pred = predict(m_svy, newdata = new, se.fit = TRUE)
pred

```

## Practice exercises

1. Replace `country_name` with a different country and compare the weighted and unweighted correlations.
2. Fit the survey-weighted model for one country and interpret the slope in a single sentence.
3. Change the prediction grade from 3 to 6. How does the predicted score change?
4. Repeat the analysis using `ch02` (age) as the predictor instead of grade. Compare the interpretation with the grade model.
